Adding selection to \stdpopsim: oh how the tables have turned

### 2024-12-19 analysis2_manuscript.tex

    \begin{figure}[t]
        \centering
        \includegraphics[width=\textwidth]{figures/HomSap/OOA/estimated_Ne_t_final}
        \caption{
        \label{fig:1pop-human-demography}
        Performance of methods to infer $N_e(t)$ from a human out-of-Africa model \citep{ragsdale2019models}
        with and without purifying selection on exons. The left column shows estimates of $N_e(t)$ from simulations
        without selection, while the right column shows estimates of $N_e(t)$ from simulations with a gamma-distributed   
        DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
        \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
        \rngcomment{Why do some methods have confidence intervals (like stairwayplot) and others don't (like msmc)?}
        }
    \end{figure}
    %DRS-fig: reminder to update the legend to just say ``exons'' instead of ensembl_havana_whatever...

@@

    \begin{figure}[htbp]
        \centering
        \includegraphics[width=\linewidth]{figures/HomSap/OOA/HomSap_OutOfAfricaArchaicAdmixture_5R19_Gamma_K17_ensembl_havana_104_exons_DFE_plot}
        \caption{A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
        Simulations were performed using a human out-of-Africa model \citep{ragsdale2019models} with a gamma-distributed DFE
        acting on exons parameterized by a mean selection coefficient and a shape parameter. Estimates of the 
        parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
        The human model has three extant populations (CEU, CHB, and YRI), and parameter estimates from each
        sample are shown in different colors.
        Estimates from three different methods are shown: 1) \grapes \cite{galtier2016adaptive}, 2) \polydfe \citep{tataru2020polydfe},
        and 3) \dadicli \citep{Huang2023}.
            \rngcomment{Update figure legend. dadi=> dadi-cli.}
}
        \label{fig:homsap-dfe.ooa}
    \end{figure}

    In Figures \ref{fig:homsap-dfe.constant} and \ref{fig:homsap-dfe.ooa} we show a comparison of estimates
    from three software packages for inferring the DFE from genetic data:
    \dadicli \citep{Huang2023}, \polydfe \citep{tataru2020polydfe}, 
    and \grapes \citep{galtier2016adaptive}.
    In each case we note that the DFE is estimated from segregating sites only,
    without the use of substitution data, as a current limitation of \stdpopsim
    is that we are not able to simulate an outgroup sequence to estimate
    divergence. 
    Figure \ref{fig:homsap-dfe.constant} shows estimates from a single, constant size population of humans, 
    while Figure \ref{fig:homsap-dfe.ooa} shows estimates from a model of human out-of-Africa demography.
    In a constant size population, all methods perform reasonably well, although \dadicli seems to
    perform slightly worse than the other methods examined, at least for the small number of replicates we simulated here.
    Notably only \polydfe overlapped the true mean selection coefficient, whereas 
    the other methods do not. 
    For the shape parameter in constant size simulations \grapes and \polydfe are roughly
    equivalent in their performance whereas \dadicli seems to yield overestimates. 
    \rngcomment{Has Fig.~\ref{fig:homsap-dfe.constant} been updated for the new GRAPES scaling?}

@@

    %add PhoSin DFE figure here
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{figures/PhoSin/Vaquita2Epoch_1R22/PhoSin_Vaquita2Epoch_1R22_Gamma_R22_Phocoena_sinus.mPhoSin1.pri.110_exons_DFE_plot.pdf}
        \caption{
        \label{fig:vaquita-dfe}
        A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
        Simulations were performed using a model of vaquita porpoise demography \citep{robinson2022critically} with a gamma-distributed DFE
        acting on exons parameterized by a mean selection coefficient, a shape parameter, and a relationship between selection and dominance. Estimates of the 
        parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
        This vaquita model has a single extant population, and parameter estimates from each
        of the three different methods are shown: 1) \grapes \cite{galtier2016adaptive}, 2) \polydfe \citep{tataru2020polydfe},
        and 3) \dadicli \citep{Huang2023,kim2017inference}.
        C) Shown is the simulated distribution of $2 h s$, where $h$ is the dominance coefficient, along with the simulated distribution of $s$ and the median inferred DFEs. The distribution of $2 h s$ is multimodal because the simulated relationship between $h$ and $s$ is not continuous.
            \rngcomment{Update figure legend. dadi=> dadi-cli.}
        }
    \end{figure}

@@

% constant size DFE figure
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/HomSap/Constant/HomSap_Constant_Gamma_K17_ensembl_havana_104_exons_DFE_plot}
    \caption{
    Performance of methods to infer the distribution of fitness effects (DFE) from genetic data.
    Simulations were performed using a human constant size model with a gamma-distributed DFE
    acting on exons parameterized by a mean selection coefficient and a shape parameter. On each row we show the true parameter 
    as a black line, and the estimates as boxplots from replicate simulations. The human model has three extant populations (CEU, CHB, and YRI).
    %DRS-fig: only one population is shown here, but 3 referenced in the legend. Wrong legend or wrong fig?
    %DRS-fig: If it is the right fig, get rid of the pop0 legend inset
    \rngcomment{Update figure legend. dadi=> dadi-cli.}
    Each row of the figure shows estimates from a different method: 1) GRAPES \citep{galtier2016adaptive}, 2) polyDFE \citep{tataru2020polydfe}, and 3) \dadicli \citep{Huang2023}.
    }
    \label{fig:homsap-dfe.constant}    
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/PhoSin/Constant/PhoSin_Constant_Gamma_R22_Phocoena_sinus.mPhoSin1.pri.110_exons_DFE_plot.pdf}
    \caption{
    \label{fig:vaquita-dfe.constant}
    A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
    Simulations were performed using a model of vaquita porpoise genome under constant population size with a gamma-distributed DFE
    acting on exons parameterized by a mean selection coefficient and a shape parameter. Estimates of the 
    parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
    This vaquita model has a single extant population, and parameter estimates from each
    of the three different methods are shown: 1) GRAPES \citep{galtier2016adaptive}, 2) polyDFE \citep{tataru2020polydfe},
    and 3) \dadicli \citep{Huang2023}.}
    \rngcomment{Update figure legend. dadi=> dadi-cli.}
    %DRS-fig: remove the pop 0 legend again
\end{figure}
% DRS-fig: ditch the unnecessary legends from panels A and B. Also, are we deciding which version of panel C to include here? I guess I prefer
% to bottom one but with the legend of the top one (``Original s DFE'' sounds a little clunky).

@@

\begin{table}[ht]
\centering
\small
\caption{\bf{Performance of DFE methods across the simulation scenarios}. 
Bold rows show the lowest mean absolute error (MAE) for the two DFE parameters
for each species.}
\begin{tabular}{lllrrrrrr}
\toprule
species ID & demography & annotation & \makecell{MAE \\ $E|s|$ \\ dadi} & \makecell{MAE \\ $E|s|$ \\ grapes} & \makecell{MAE \\ $E|s|$ \\ polyDFE} & \makecell{MAE \\ shape \\ dadi} & \makecell{MAE \\ shape \\ grapes} & \makecell{MAE \\ shape \\ polyDFE} \\
\midrule
HomSap & Constant & all sites & 0.0027 & 0.0111 & \bf{0.0018} & \bf{0.00071} & 0.029 & 0.014 \\
HomSap & Constant & exons & \bf{0.0012} & 0.0101 & 0.0023 & 0.030 & 0.0086 & \bf{0.0068} \\
HomSap & OOA 5R19 & all sites & 0.012 & 0.012 & \bf{0.0074} & \bf{0.027} & 0.055 & 0.035 \\
HomSap & OOA 5R19 & exons & 0.014 & 0.0111 & \bf{0.0072} & 0.031 & 0.051 & \bf{0.024} \\
PhoSin & Constant & all sites & \bf{0.024} & 0.025 & 0.024 & 0.21 & 0.24 & \bf{0.19} \\
PhoSin & Constant & exons & 0.024 & 0.025 & 0.024 & \bf{0.23} & 0.25 & 0.23 \\
PhoSin & Vaquita2Epoch 1R22 & all sites & \bf{0.024} & 0.025 & 0.024 & 0.20 & 0.23 & \bf{0.18} \\
PhoSin & Vaquita2Epoch 1R22 & exons & 0.024 & 0.025 & \bf{0.024} & \bf{0.18} & 0.23 & 0.21 \\
\bottomrule
\end{tabular}
\label{tab:dfe_table}
\end{table}

>>>>

    \begin{figure}[t]
        \centering
        \includegraphics[width=\textwidth]{figures/HomSap/OOA/estimated_Ne_t_final}
        \caption{
        \label{fig:1pop-human-demography}
        Performance of methods to infer $N_e(t)$ from a human out-of-Africa model \citep{ragsdale2019models}
        with and without purifying selection on exons. The left column shows estimates of $N_e(t)$ from simulations
        without selection, while the right column shows estimates of $N_e(t)$ from simulations with a gamma-distributed   
        DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
        \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
        }
    \end{figure}
    %DRS-fig: reminder to update the legend to just say ``exons'' instead of ensembl_havana_whatever...

@@

    \begin{figure}[htbp]
        \centering
        \includegraphics[width=\linewidth]{figures/HomSap/OOA/HomSap_OutOfAfricaArchaicAdmixture_5R19_Gamma_K17_ensembl_havana_104_exons_DFE_plot}
        \caption{A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
        Simulations were performed using a human out-of-Africa model \citep{ragsdale2019models} with a gamma-distributed DFE
        acting on exons parameterized by a mean selection coefficient and a shape parameter. Estimates of the 
        parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
        The human model has three extant populations (CEU, CHB, and YRI), and parameter estimates from each
        sample are shown in different colors.
        Estimates from three different methods are shown: 1) \grapes \cite{galtier2016adaptive}, 2) \polydfe \citep{tataru2020polydfe},
        and 3) \dadicli \citep{Huang2023}.
}
        \label{fig:homsap-dfe.ooa}
    \end{figure}

    In Figures \ref{fig:homsap-dfe.constant} and \ref{fig:homsap-dfe.ooa} we show a comparison of estimates
    from three software packages for inferring the DFE from genetic data:
    \dadicli \citep{Huang2023}, \polydfe \citep{tataru2020polydfe}, 
    and \grapes \citep{galtier2016adaptive}.
    In each case we note that the DFE is estimated from segregating sites only,
    without the use of substitution data, as a current limitation of \stdpopsim
    is that we are not able to simulate an outgroup sequence to estimate
    divergence. 
    Figure \ref{fig:homsap-dfe.constant} shows estimates from a single, constant size population of humans, 
    while Figure \ref{fig:homsap-dfe.ooa} shows estimates from a model of human out-of-Africa demography.
    In a constant size population, all methods perform reasonably well, although \dadicli seems to
    perform slightly worse than the other methods examined, at least for the small number of replicates we simulated here.
    Notably only \polydfe overlapped the true mean selection coefficient, whereas 
    the other methods do not. 
    For the shape parameter in constant size simulations \grapes and \polydfe are roughly
    equivalent in their performance whereas \dadicli seems to yield overestimates.

@@

    %add PhoSin DFE figure here
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{figures/PhoSin/Vaquita2Epoch_1R22/PhoSin_Vaquita2Epoch_1R22_Gamma_R22_Phocoena_sinus.mPhoSin1.pri.110_exons_DFE_plot.pdf}
        \caption{
        \label{fig:vaquita-dfe}
        A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
        Simulations were performed using a model of vaquita porpoise demography \citep{robinson2022critically} with a gamma-distributed DFE
        acting on exons parameterized by a mean selection coefficient, a shape parameter, and a relationship between selection and dominance. Estimates of the 
        parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
        This vaquita model has a single extant population, and parameter estimates from each
        of the three different methods are shown: 1) \grapes \cite{galtier2016adaptive}, 2) \polydfe \citep{tataru2020polydfe},
        and 3) \dadicli \citep{Huang2023,kim2017inference}.
        C) Shown is the simulated distribution of $2 h s$, where $h$ is the dominance coefficient, along with the simulated distribution of $s$ and the median inferred DFEs. The distribution of $2 h s$ is multimodal because the simulated relationship between $h$ and $s$ is not continuous.
        }
    \end{figure}

@@

% constant size DFE figure
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/HomSap/Constant/HomSap_Constant_Gamma_K17_ensembl_havana_104_exons_DFE_plot}
    \caption{
    Performance of methods to infer the distribution of fitness effects (DFE) from genetic data.
    Simulations were performed using a human constant size model with a gamma-distributed DFE
    acting on exons parameterized by a mean selection coefficient and a shape parameter. On each row we show the true parameter 
    as a black line, and the estimates as boxplots from replicate simulations. The human model has three extant populations (CEU, CHB, and YRI).
    %DRS-fig: only one population is shown here, but 3 referenced in the legend. Wrong legend or wrong fig?
    %DRS-fig: If it is the right fig, get rid of the pop0 legend inset
    Each row of the figure shows estimates from a different method: 1) GRAPES \citep{galtier2016adaptive}, 2) polyDFE \citep{tataru2020polydfe}, and 3) \dadicli \citep{Huang2023}.
    }
    \label{fig:homsap-dfe.constant}    
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/PhoSin/Constant/PhoSin_Constant_Gamma_R22_Phocoena_sinus.mPhoSin1.pri.110_exons_DFE_plot.pdf}
    \caption{
    \label{fig:vaquita-dfe.constant}
    A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
    Simulations were performed using a model of vaquita porpoise genome under constant population size with a gamma-distributed DFE
    acting on exons parameterized by a mean selection coefficient and a shape parameter. Estimates of the 
    parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
    This vaquita model has a single extant population, and parameter estimates from each
    of the three different methods are shown: 1) GRAPES \citep{galtier2016adaptive}, 2) polyDFE \citep{tataru2020polydfe},
    and 3) \dadicli \citep{Huang2023}.}
    %DRS-fig: remove the pop 0 legend again
\end{figure}
% DRS-fig: ditch the unnecessary legends from panels A and B. Also, are we deciding which version of panel C to include here? I guess I prefer
% to bottom one but with the legend of the top one (``Original s DFE'' sounds a little clunky).

@@

\begin{table}[ht]
\centering
\small
\caption{\bf{Performance of DFE methods across the simulation scenarios}. 
Bold rows show the lowest mean absolute error (MAE) for the two DFE parameters
for each species.}
\begin{tabular}{lllrrrrrr}
\toprule
species ID & demography & annotation & \makecell{MAE \\ $E|s|$ \\ dadi-cli} & \makecell{MAE \\ $E|s|$ \\ grapes} & \makecell{MAE \\ $E|s|$ \\ polyDFE} & \makecell{MAE \\ shape \\ dadi-cli} & \makecell{MAE \\ shape \\ grapes} & \makecell{MAE \\ shape \\ polyDFE} \\
\midrule
HomSap & Constant & all sites & 0.0027 & 0.0111 & \bf{0.0018} & \bf{0.00071} & 0.029 & 0.014 \\
HomSap & Constant & exons & \bf{0.0012} & 0.0101 & 0.0023 & 0.030 & 0.0086 & \bf{0.0068} \\
HomSap & OOA 5R19 & all sites & 0.012 & 0.012 & \bf{0.0074} & \bf{0.027} & 0.055 & 0.035 \\
HomSap & OOA 5R19 & exons & 0.014 & 0.0111 & \bf{0.0072} & 0.031 & 0.051 & \bf{0.024} \\
PhoSin & Constant & all sites & \bf{0.024} & 0.025 & 0.024 & 0.21 & 0.24 & \bf{0.19} \\
PhoSin & Constant & exons & 0.024 & 0.025 & 0.024 & \bf{0.23} & 0.25 & 0.23 \\
PhoSin & Vaquita2Epoch 1R22 & all sites & \bf{0.024} & 0.025 & 0.024 & 0.20 & 0.23 & \bf{0.18} \\
PhoSin & Vaquita2Epoch 1R22 & exons & 0.024 & 0.025 & \bf{0.024} & \bf{0.18} & 0.23 & 0.21 \\
\bottomrule
\end{tabular}
\label{tab:dfe_table}
\end{table}
